---
title: "Citibike User Behavior"
author: "Corentin Llorca (cl3783)"
date: "24 November 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE,
                      echo = FALSE)
```


```{r}
library(tidyverse)
data <- read_csv("../data/2018-feb-jul-oct-citibike-trips.csv")
```

## Who uses Citi Bike ?


### A brief demographic study of Citi Bike users : age and gender

The Citi Bike public database contains informations about the age and gender of the users. However, that is only the case for users with the subscriptions, as simple customers with short-term passes aren't required to fill in any information. Still, we have seen that subscriber trips represented about 90% of all trips, so the following studies on age and gender can be safely extended to the whole user base of Citi Bike.


### Customers and Subscribers, and their usage of Citi Bike

Here, we will be looking at the "Customer Type" variable in the trip database. It is an unordered factor corresponding to the type of customer having taken the trip. The different levels are "Customer" and "Subscriber". The Customers are defined as the people using a 1 or 3 day pass, thus more likely to be occasional users, and the Subscribers are those with the annual pass, thus more likely to be regular users. 

The two levels are highly unbalanced, as we can see below : 

```{r}
dplyr::count(data,usertype)
```

We would expect a service like Citi Bike to be used by subscribers much more than by occasional users, since the annual subscription (170 dollars) is much more cost-efficient than the 1-day pass (12 dollars). We observe a roughly 10%/90% repartition. However, the size of our data (more than 4.5 million trips) makes it so that we can still safely assume that the trends observed among customers are actually meaningful.

#### Seasonal patterns 

```{r}

p4 <- ggplot(data,aes(x = format(data$starttime,"%B"), fill = usertype)) + 
      geom_bar(color = 'black') + 
      ggtitle("Number of trips per user type for each month") + 
      xlab("Month") + ylab("Count") + 
      scale_fill_discrete(name = "User type")
p4
```

First of all, this confirms that an overwhelming majority of trips are taken by subscribers with the annual pass. As for the variation in the repartition between customers and subscribers, we can clearly see an evolution. We can see that there is a much lower proportion of customers compared to subscribers in the winter than in the summer. Moreover, between July and October, we see the number of customers decrease while the number of subscribers increases: this is a clear indicator that there is a lower proportion of customers in October than in July, although that proportion is still much higher than in February. 

This seems logical, as we would expect occasional users to use Citi Bike when the weather is more forgiving. Also, we might emit the hypothesis that Citi Bike is used by tourists - and tourists would in all likelihood be customers rather than subscribers. Then, the increasing number of tourists in the summer compared to the winter might be another cause for that observed difference. 


#### Weekly patterns 

We've already seen that Citi Bikes were being used less on weekends. Let's now see whether the weekly usage differs between the two categories of customers.

```{r}

p5 <- ggplot(data,aes(x = format(data$starttime,"%A"), fill = usertype)) + 
      geom_bar(color = 'black') + 
      ggtitle("Number of trips per user type for each day of the week") + 
      xlab("Weekday") + ylab("Count") + 
      scale_x_discrete (limits = 
          c("Monday","Tuesday","Wednesday","Thursday","Friday","Saturday","Sunday"))+
      scale_fill_discrete(name = "User type")
p5
```

We see here that customers actually use Citi Bike on weekends more than they do during the week, while it's the opposite for subscribers. This suggests that subscribers might use Citi Bike to commute to work, while customers use it more for leisure, hence the important use on weekends.


#### Daily patterns


```{r}

p6 <- ggplot(data,
             aes(x = as.numeric(format(data$starttime,"%H")) +
                     as.numeric(format(data$starttime,"%M"))/60 +
                     as.numeric(format(data$starttime,"%S"))/3600,
                 color = usertype)) + 
      geom_density() +  
      ggtitle("Repartition of trip times for each user type") + 
      xlab("Hour of the day") + ylab("Density") + 
      scale_x_continuous(breaks = seq(0,24,4)) +
      scale_color_discrete(name = "User type")
p6
```

We notice a clear difference in the density curves between customers and subscribers, and this difference is another strong piece of evidence in favor of the hypothesis that customers might be in majority tourists. Indeed, we see a curve that's roughly bi-modal for subscribers, with modes aroung 8 a.m. and 6 p.m., which are times usually associated with commuting. On the other hand, the curve for customers is unimodal, with the mode roughly around 3 p.m., and a much slower start to the day; which indicates that the bikes are being used for leisure. 


#### Types of trips taken

##### Trip duration

We want to highlight the difference in the types of trips taken by customers and subscribers. First of all, we want to see if there's any difference in the duration of those trips.

Here, we plot vertical lines for both plots that signal the end of the time included in the plan, which is 30 minutes for customers and 45 for subscribers. 

```{r}
line_data = tibble(usertype = c("Customer","Subscriber"), vline = c(30,45))

breaks_21 = seq(0,120,3)
p7 <- ggplot(data,aes(x = tripduration/60)) + 
       geom_histogram(aes(y = ..density..),
                      breaks = breaks_21, 
                      fill = "lightblue",color = "black") + 
       facet_wrap(~usertype) + 
       geom_vline(data = line_data, aes(xintercept = vline)) +
       ggtitle("Repartition of trip durations for each user type") + 
       xlab("Trip duration (minutes)") + ylab("Density")
p7
```

Once again, we can notice a clear difference between both repartitions. Indeed, subscribers tend to favor shorter trips, with the overwhelming majority of trips taken being around 10 to 15 minutes long - that might correspond to the time for commuting. Customers, on the other hand, take longer trips, with the mode being around 25 minutes. This can be explained by the fact that we consider some customers to be tourists, and tourists would explore the city much more than residents, thus taking longer-lasting trips.

Moreover, we can notice that customers are generally much keener on taking long trips, going as far as to make them longer than the time that is included in the plan. Indeed, even though we do notice a significant dropoff between the 25-30 and 30-35 minutes bins, a significant part of customer trips are still longer than 30 minutes, with some of them going well past 1 hour. As for subscribers, they usually take very few trips longer than 45 minutes, which is their alloted time limit.

Something that might also reveal something would be the average trip length for each hour of the day. Let's show that plot : 

```{r}

p8 <- ggplot(data,
             aes(x = factor(
                  as.numeric(format(data$starttime,"%H"))),
                 y = tripduration,
                 color = usertype)) + 
      stat_summary(fun.y = 'mean', geom = "point") +  
      ggtitle("Average trip duration per hour of day for each user type") + 
      xlab("Hour of the day") + ylab("Trip duration") + 
      scale_y_continuous(breaks = seq(0,1600,200)) +
      ylim(0,1600) +
      scale_color_discrete(name = "User type")
p8
```

First of all, we get a confirmation that the average trip is longer for customers than for subscribers. 

### A brief demographic study of Citi Bike users : age and gender

The Citi Bike public database contains informations about the age and gender of the users. However, that is only the case for users with the subscriptions, as simple customers with short-term passes aren't required to fill in any information. Still, we have seen that subscriber trips represented about 90% of all trips, so the following studies on age and gender can be safely extended to the whole user base of Citi Bike.

Let's first show some simple plots. Let's look
